---
title: "Mouse Experiment Analysis Example"
format: html
editor: source
---

```{r setup, message=FALSE, warning=FALSE}
library(mouseExperiment)
library(dplyr)
library(ggplot2)

```

## Introduction

This document demonstrates the functionality of the mouseExperiment R package for analyzing and visualizing mouse tumor growth data.

## Data Preparation

First, we'll load the synthetic dataset included with the package:

```{r load-data}
data(synthetic_data)
head(synthetic_data)
```

Now we'll calculate tumor volumes and experiment days:

```{r calculate}

source("../R/calculate_dates.R")
source("../R/calculate_volume.R")
# Calculate tumor volume
df <- calculate_volume(synthetic_data)

# Calculate experiment days from start date
df <- calculate_dates(df, start_date = "2022-02-23")

# Display the resulting data
head(df)
```

## Tumor Growth Visualization

Let's visualize the tumor growth over time by treatment group:

```{r plot-growth}
source("../R/plot_tumor_growth.R")
plot_tumor_growth(df)
```

## Tumor Growth Statistics

We can perform statistical analysis on the tumor growth data:

```{r growth-stats, message=FALSE, warning=FALSE}
source("../R/tumor_growth_statistics.R")
# Standard tumor growth statistics
growth_stats <- tumor_growth_statistics(df)

```

## Survival Analysis

Now let's analyze and visualize survival data:

```{r survival-plot}
source("../R/plot_survival.R")
# Plot survival curves by group
plot_survival(df)
```

```{r survival-stats}
source("../R/survival_statistics.R")

surv_stats = survival_statistics(df)

surv_stats$model
surv_stats$results

```

## Bayesian Analysis

The package also supports Bayesian analysis of tumor growth data:

```{r bayes-analysis, eval=FALSE}

source("../R/tumor_growth_statistics_bayes.R")

# Bayesian tumor growth statistics 
# Note: This is computationally intensive, so we're not running it in this example
bayesian_stats <- tumor_growth_statistics_bayes(df, cores = 4)
# print(bayesian_stats)

#Returning model without Bayes factor.Error in exists("log_bf") && !is.na(log_bf) : 
  'length = 2' in coercion to 'logical(1)'

```

```{r}

source("../R/plot_caterpillar.R")

# Create a basic Bayesian caterpillar plot
  bayes_cat_plot <- plot_caterpillar_bayes(bayesian_stats$model)
  print(bayes_cat_plot)

  # Show random effects (mouse-specific variations)
  bayes_cat_plot_with_ranef <- plot_caterpillar_bayes(
    bayesian_stats$model,
    show_random = TRUE,
    title = "All Model Parameters with 95% Credible Intervals"
  )
  print(bayes_cat_plot_with_ranef)

```


## Conclusion

This vignette demonstrates the key functionality of the mouseExperiment package for analyzing mouse tumor growth experiment data.
