---
title: "mouseExperiment: Comprehensive Analysis of Tumor Growth Experiments"
author: "mouseExperiment Package"
format: 
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: cosmo
    highlight-style: github
vignette: >
  %\VignetteIndexEntry{mouseExperiment: Comprehensive Analysis of Tumor Growth Experiments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8, 
  fig.height = 6,
  out.width = "100%"
)
```

# Introduction

The `mouseExperiment` package provides a comprehensive set of tools for analyzing preclinical mouse tumor growth experiments. This vignette demonstrates how to use the package's core functionality with example datasets included in the package.

## Package Installation

```{r eval=FALSE}
# Install from GitHub
devtools::install_github("username/mouseExperiment")
```

## Loading the Package

Let's start by loading the package:

```{r}
library(mouseExperiment)
```

# 1. Basic Data Processing

## 1.1 Calculating Tumor Volume

The first step in processing tumor measurement data is to calculate tumor volumes from length and width measurements. The `calculate_volume()` function supports various volume calculation formulas.

We'll use the combo treatment dataset for this demonstration:

```{r}
# Load the data
data(combo_treatment_synthetic_data)

# Display the first few rows
head(combo_treatment_synthetic_data)
```

Now, let's calculate tumor volumes using the default ellipsoid formula:

```{r}
# Calculate volume using the ellipsoid formula (default)
data_with_volume <- calculate_volume(combo_treatment_synthetic_data)

# Display first few rows with volume
head(data_with_volume)
```

The ellipsoid formula V = (length × width² × π) / 6 is the most widely used and accurate formula for ovoid tumors.

## 1.2 Converting Dates to Experimental Days

In mouse experiments, it's standard to convert calendar dates to experimental days, where day 0 is typically the day of tumor inoculation or the start of treatment. The `calculate_dates()` function handles this conversion:

```{r}
# Calculate days since the start of the experiment
data_processed <- calculate_dates(data_with_volume, 
                                 start_date = "03/24/2025", 
                                 date_format = "%m/%d/%Y")

# Display the result
head(data_processed)
```

Now the data has a "Day" column that represents the number of days since the start of the experiment.

# 2. Visualization of Tumor Growth

## 2.1 Basic Tumor Growth Plots

The `plot_tumor_growth()` function creates visualizations of tumor growth over time:

```{r}
# Create a basic tumor growth plot
plot_tumor_growth(data_processed, 
                 id_column = "ID", 
                 group_column = "Treatment",
                 facet = TRUE)
```

## 2.2 Highlighting Treatment Days

We can indicate treatment days on the tumor growth plot:

```{r}
# Plot with treatment days marked (e.g., treatments on days 0, 3, 7, 10)
plot_tumor_growth(data_processed, 
                 id_column = "ID", 
                 group_column = "Treatment",
                 treatment_days = c(0, 3, 7, 10),
                 treatment_label = "Treatment administered")
```

## 2.3 Combined Visualization

We can create a comprehensive visualization showing individual growth curves, group means, and treatment days:

```{r}
# Create a comprehensive plot
plot_tumor_growth(data_processed, 
                 id_column = "ID", 
                 group_column = "Treatment",
                 treatment_days = c(0, 3, 7, 10),
                 treatment_label = "Treatment administered",
                 show_individual = TRUE,
                 show_mean = TRUE,
                 show_error = TRUE)
```

# 3. Statistical Analysis of Tumor Growth

## 3.1 Mixed-Effects Model Analysis

The `tumor_growth_statistics()` function fits a linear mixed-effects model to analyze tumor growth data:

```{r}
# Run statistical analysis
growth_stats <- tumor_growth_statistics(data_processed, 
                                      volume_column = "Volume", 
                                      group_column = "Treatment",
                                      id_column = "ID")

# Display ANOVA table for fixed effects
growth_stats$anova
```

This analysis tests for the effects of time, treatment, and their interaction on tumor growth.

## 3.2 Visualizing the Model Coefficients

We can visualize the model coefficients using a caterpillar plot:

```{r}
# Plot model coefficients with confidence intervals
plot_caterpillar(growth_stats$model)
```

## 3.3 Post-hoc Comparisons

The function also provides post-hoc pairwise comparisons between treatment groups:

```{r}
# Display post-hoc comparisons
growth_stats$posthoc
```

# 4. Survival Analysis

## 4.1 Survival Plots

The `plot_survival()` function creates Kaplan-Meier survival curves:

```{r}
# Create survival plot
survival_plot <- plot_survival(data_processed, 
                             time_column = "Day", 
                             censor_column = "Survival_Censor",
                             treatment_column = "Treatment",
                             show_legend = TRUE,
                             show_pvalue = TRUE)

# Display the plot
print(survival_plot)
```

## 4.2 Survival Statistics

The `survival_statistics()` function performs a Cox proportional hazards analysis:

```{r}
# Perform survival analysis
survival_results <- survival_statistics(data_processed, 
                                      time_column = "Day",
                                      censor_column = "Survival_Censor",
                                      treatment_column = "Treatment",
                                      reference_group = "Control")

# Display hazard ratios and confidence intervals
survival_results$results
```

## 4.3 Forest Plot of Hazard Ratios

The function also creates a forest plot to visualize hazard ratios:

```{r}
# Display forest plot
survival_results$forest_plot
```

# 5. Dose-Response Analysis

The package includes tools for analyzing dose-response relationships. Let's use the dose levels dataset:

```{r}
# Load dose-response data
data(dose_levels_synthetic_data)

# Process the data
dose_data <- calculate_volume(dose_levels_synthetic_data)
dose_data <- calculate_dates(dose_data, 
                           start_date = "24-Mar", 
                           date_format = "%d-%b", 
                           year = 2025)

# Analyze dose-response relationship
dose_results <- dose_response_statistics(dose_data, 
                                       dose_column = "Dose",
                                       treatment_column = "Treatment",
                                       time_point = 14)  # Analyze at day 14

# Display analysis summary
dose_results$summary_table
```

## 5.1 Dose-Response Visualization

The function creates several visualizations:

```{r}
# Display scatter plot with regression line
dose_results$plots$scatter
```

```{r}
# Display box plot by dose level
dose_results$plots$boxplot
```

```{r}
# Display bar plot with error bars
dose_results$plots$barplot
```

# 6. Drug Synergy Analysis

The `analyze_drug_synergy()` function tests for synergistic effects between two treatments:

```{r}
# Analyze synergy between HDACi and PD1 treatments
synergy_results <- analyze_drug_synergy(data_processed, 
                                      time_point = 21,
                                      treatment_groups = c("Control", "HDACi", "aPD1", "HDACi + PD1"))

# Display synergy test results
synergy_results$synergy_test
```

## 6.1 Synergy Visualization

The function also creates visualizations of the observed vs. expected effects:

```{r}
# Display bar plot of observed vs. expected effects
synergy_results$plots$barplot
```

# 7. Bayesian Analysis

For more complex modeling needs, the package provides Bayesian analysis with the `tumor_growth_statistics_bayes()` function:

```{r eval=FALSE}
# Perform Bayesian analysis
# Note: This can take several minutes to run
bayes_results <- tumor_growth_statistics_bayes(data_processed, 
                                             volume_column = "Volume", 
                                             group_column = "Treatment",
                                             id_column = "ID",
                                             cores = 4)  # Using 4 cores for parallel processing

# Visualize posteriors
plot_caterpillar_bayes(bayes_results$model)
```

# 8. Power Analysis

The `post_power_analysis()` function helps calculate achieved power based on actual results:

```{r}
# Perform post-hoc power analysis
power_results <- post_power_analysis(data_processed, 
                                   volume_column = "Volume", 
                                   group_column = "Treatment",
                                   id_column = "ID",
                                   time_column = "Day",
                                   time_point = 21)

# Display power analysis results
power_results$power_table
```

# Summary

The `mouseExperiment` package provides a comprehensive toolset for analyzing mouse tumor growth experiments:

1. **Data Processing**: Calculate volumes and experimental days
2. **Visualization**: Create informative plots of tumor growth and survival
3. **Statistical Analysis**: Perform mixed-effects modeling and survival analysis
4. **Dose-Response Analysis**: Test for dose-response relationships
5. **Drug Synergy Analysis**: Evaluate combination treatments
6. **Bayesian Analysis**: Perform more complex modeling
7. **Power Analysis**: Calculate achieved statistical power

These tools help researchers extract meaningful insights from preclinical mouse experiments and support rigorous statistical analysis of the data.