---
title: "Analyzing Mouse Tumor Growth Experiments with mouseExperiment"
author: "David Woods"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyzing Mouse Tumor Growth Experiments with mouseExperiment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)
```

# Introduction

The `mouseExperiment` package provides a comprehensive suite of tools for analyzing mouse tumor growth experiments. This vignette demonstrates the main functionality of the package, including:

- Tumor growth analysis using mixed-effects models
- Survival analysis with Kaplan-Meier curves
- Synergy analysis for combination treatments
- Visualization tools for various analyses

# Installation

```{r install, eval=FALSE}
# Install from CRAN
install.packages("mouseExperiment")

# Or install the development version from GitHub
# devtools::install_github("yourusername/mouseExperiment")
```

# Loading the Package

```{r load}
library(mouseExperiment)
```

# Tumor Growth Analysis

## Basic Usage

The `tumor_growth_statistics()` function analyzes tumor growth data using either linear mixed-effects models (lme4) or area under the curve (AUC) analysis.

```{r tumor_growth}
# Load example data
data(combo_treatment_synthetic_data)

# Analyze tumor growth using mixed-effects model
results <- tumor_growth_statistics(
  data = combo_treatment_synthetic_data,
  time_column = "Day",
  volume_column = "Volume",
  id_column = "Mouse_ID",
  treatment_column = "Treatment",
  cage_column = "Cage",
  model_type = "lme4"
)

# Print summary of results
print_results(results)
```

## Visualizing Growth Rates

The `plot_growth_rate()` function creates visualizations of tumor growth rates:

```{r plot_growth}
# Create growth rate plot
p <- plot_growth_rate(
  results$growth_rates,
  treatment_column = "Treatment",
  growth_rate_column = "growth_rate",
  title = "Tumor Growth Rates by Treatment",
  ylab = "Growth Rate (mm³/day)"
)
print(p)
```

## Area Under the Curve Analysis

You can also analyze tumor growth using AUC:

```{r auc_analysis}
# Perform AUC analysis
auc_results <- tumor_growth_statistics(
  data = combo_treatment_synthetic_data,
  time_column = "Day",
  volume_column = "Volume",
  id_column = "Mouse_ID",
  treatment_column = "Treatment",
  cage_column = "Cage",
  model_type = "auc"
)

# Visualize AUC results
p_auc <- plot_auc(
  auc_results$auc_data,
  group_column = "Treatment",
  auc_column = "AUC",
  title = "Area Under the Curve by Treatment",
  ylab = "AUC (mm³·days)"
)
print(p_auc)
```

# Survival Analysis

## Basic Usage

The `survival_statistics()` function performs survival analysis:

```{r survival}
# Load example survival data
data(survival_synthetic_data)

# Perform survival analysis
surv_results <- survival_statistics(
  data = survival_synthetic_data,
  time_column = "Time",
  event_column = "Event",
  treatment_column = "Treatment",
  id_column = "Mouse_ID"
)

# Print summary
print_results(surv_results)
```

## Visualizing Survival Curves

The `plot_survival()` function creates Kaplan-Meier survival curves:

```{r plot_survival}
# Create survival plot
p_surv <- plot_survival(
  surv_results$survival_data,
  time_column = "Time",
  event_column = "Event",
  treatment_column = "Treatment",
  title = "Survival Curves by Treatment",
  risk_table = TRUE
)
print(p_surv)
```

## Forest Plot

The `forest_plot()` function visualizes hazard ratios:

```{r forest_plot}
# Create forest plot
p_forest <- forest_plot(
  surv_results$summary,
  title = "Hazard Ratios by Treatment"
)
print(p_forest)
```

# Synergy Analysis

## Bliss Independence

The `bliss_independence()` function calculates Bliss synergy:

```{r bliss}
# Calculate Bliss synergy
bliss_results <- bliss_independence(
  data = combo_treatment_synthetic_data,
  time_column = "Day",
  volume_column = "Volume",
  id_column = "Mouse_ID",
  treatment_column = "Treatment",
  cage_column = "Cage"
)

# Visualize Bliss synergy
p_bliss <- plot_bliss(
  bliss_results,
  title = "Bliss Synergy Over Time",
  ylab = "Bliss Synergy Score"
)
print(p_bliss)
```

## Combination Index

The `synergy_stats()` function calculates the Combination Index:

```{r synergy}
# Calculate Combination Index
synergy_results <- synergy_stats(
  data = combo_treatment_synthetic_data,
  time_column = "Day",
  volume_column = "Volume",
  id_column = "Mouse_ID",
  treatment_column = "Treatment",
  cage_column = "Cage"
)

# Visualize Combination Index
p_ci <- plot_combination_index(
  synergy_results$synergy_summary,
  title = "Combination Index Over Time",
  ylab = "Combination Index"
)
print(p_ci)
```

# Additional Features

## Customizing Plots

All plotting functions support customization:

```{r custom_plot}
# Customize growth rate plot
p_custom <- plot_growth_rate(
  results$growth_rates,
  treatment_column = "Treatment",
  growth_rate_column = "growth_rate",
  title = "Customized Growth Rate Plot",
  ylab = "Growth Rate (mm³/day)",
  line_color = "blue",
  point_size = 3,
  theme = theme_bw()
)
print(p_custom)
```

## Exporting Results

You can export results to various formats:

```{r export}
# Export results to CSV
write.csv(results$summary, "tumor_growth_summary.csv", row.names = FALSE)

# Save plots
ggsave("growth_rate_plot.pdf", p, width = 10, height = 6)
ggsave("survival_plot.pdf", p_surv, width = 10, height = 6)
```

# Conclusion

The `mouseExperiment` package provides a comprehensive set of tools for analyzing mouse tumor growth experiments. This vignette demonstrates the main functionality, but there are many more features available. For more information, see the package documentation and examples. 