---
title: "Analyzing Mouse Tumor Growth Experiments with mouseExperiment"
author: "David Woods"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyzing Mouse Tumor Growth Experiments with mouseExperiment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

The `mouseExperiment` package provides a comprehensive set of tools for analyzing mouse tumor growth experiments. It includes functions for:

* Calculating tumor volumes
* Analyzing growth rates
* Generating summary statistics
* Performing survival analysis
* Assessing treatment synergy
* Visualizing results

# Installation

You can install the `mouseExperiment` package from GitHub:

```{r install, eval = FALSE}
# Install devtools if not already installed
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}

# Install mouseExperiment from GitHub
devtools::install_github("yourusername/mouseExperiment")
```

# Getting Started

## Loading the Package

```{r library}
library(mouseExperiment)
```

## Available Datasets

The package includes synthetic datasets for demonstration:

```{r datasets}
data(combo_treatment_synthetic_data)
data(survival_synthetic_data)
data(single_treatment_synthetic_data)
```

# Basic Analysis

## Volume Calculation

The `calculate_volume` function converts tumor measurements to volume:

```{r volume}
# Calculate volume from dimensions
data_with_volume <- calculate_volume(
  df = combo_treatment_synthetic_data,
  length_column = "Length",
  width_column = "Width"
)

# View the first few rows
head(data_with_volume)
```

## Date Calculation

The `calculate_dates` function converts study days to actual dates:

```{r dates}
# Calculate dates from days
data_with_dates <- calculate_dates(
  df = data_with_volume,
  start_date = "01/15/2023"
)

# View the first few rows
head(data_with_dates)
```

# Tumor Growth Analysis

## Statistical Analysis

The `tumor_growth_statistics` function analyzes tumor growth:

```{r growth_stats}
# Perform statistical analysis
results <- tumor_growth_statistics(
  df = data_with_volume,
  time_column = "Day",
  volume_column = "Volume",
  treatment_column = "Treatment",
  id_column = "Mouse_ID",
  cage_column = "Cage"
)

# View summary statistics
print(results$summary)
```

## Growth Rate Visualization

The `plot_growth_rate` function visualizes tumor growth rates:

```{r growth_rate_plot}
# Create growth rate plot
p <- plot_growth_rate(
  results$growth_rates,
  treatment_column = "Treatment",
  growth_rate_column = "growth_rate",
  title = "Tumor Growth Rate by Treatment"
)
print(p)
```

## AUC Analysis

The `tumor_auc_analysis` function calculates the area under the curve:

```{r auc}
# Calculate AUC
auc_results <- tumor_auc_analysis(
  df = data_with_volume,
  time_column = "Day",
  volume_column = "Volume",
  treatment_column = "Treatment",
  id_column = "Mouse_ID"
)

# Visualize AUC
p_auc <- plot_auc(
  auc_results$auc_data,
  treatment_column = "Treatment",
  auc_column = "AUC",
  title = "Area Under the Curve by Treatment"
)
print(p_auc)
```

# Survival Analysis

The `survival_statistics` function performs survival analysis:

```{r survival}
# Perform survival analysis
surv_results <- survival_statistics(
  df = survival_synthetic_data,
  time_column = "Time",
  censor_column = "Event",
  treatment_column = "Treatment",
  id_column = "Mouse_ID"
)

# Print summary
print_results(surv_results)
```

# Synergy Analysis

## Bliss Independence

The `bliss_independence()` function calculates Bliss synergy:

```{r bliss}
# Calculate Bliss synergy
bliss_results <- bliss_independence(
  data = combo_treatment_synthetic_data,
  time_column = "Day",
  volume_column = "Volume",
  id_column = "Mouse_ID",
  treatment_column = "Treatment",
  cage_column = "Cage"
)

# Visualize Bliss synergy
p_bliss <- plot_bliss(
  bliss_results,
  title = "Bliss Synergy Over Time",
  ylab = "Bliss Synergy Score"
)
print(p_bliss)
```

## Combination Index

The `synergy_stats()` function calculates the Combination Index:

```{r synergy}
# Calculate Combination Index
synergy_results <- synergy_stats(
  data = combo_treatment_synthetic_data,
  time_column = "Day",
  volume_column = "Volume",
  id_column = "Mouse_ID",
  treatment_column = "Treatment",
  cage_column = "Cage"
)

# Visualize Combination Index
p_ci <- plot_combination_index(
  synergy_results$synergy_summary,
  title = "Combination Index Over Time",
  ylab = "Combination Index"
)
print(p_ci)
```

# Additional Features

## Customizing Plots

All plotting functions support customization:

```{r custom_plot}
# Customize growth rate plot
p_custom <- plot_growth_rate(
  results$growth_rates,
  treatment_column = "Treatment",
  growth_rate_column = "growth_rate",
  title = "Customized Growth Rate Plot",
  ylab = "Growth Rate (mm³/day)",
  line_color = "blue",
  point_size = 3,
  theme = theme_bw()
)
print(p_custom)
```

## Exporting Results

You can export results to various formats:

```{r export}
# Export results to CSV
write.csv(results$summary, "tumor_growth_summary.csv", row.names = FALSE)

# Save plots
ggsave("growth_rate_plot.pdf", p, width = 10, height = 6)
```

# Conclusion

The `mouseExperiment` package provides a comprehensive set of tools for analyzing mouse tumor growth experiments. This vignette demonstrates the main functionality, but there are many more features available. For more information, see the package documentation and examples. 