---
title: "Demo Analysis of Synthetic Combination Treatment Mouse Data"
author: "Insight BioAnalytics"
format: 
  html:
    output-dir: "../outputs/reports"
    toc: true
    toc-location: left
    number-sections: true
    embed-resources: true
    theme: lux
    logo: "/Users/david/Library/Mobile Documents/com~apple~CloudDocs/InsightBioAnalytics/logos/lightbulb/Color.png"
---

<u>Goals:</u>  
-Primary: Contrast the impact of treatments (HDAC inhibitor, PD1 blocking antibody) on tumor growth in a longitudinal murine experiment.   
-Secondary: Determine if the combination of the two treatments, an HDAC inhibitor (HDACi) and PD1 blocking antibody, is synergistic.  
  
<u>Data:</u>    
-Longitudinal tumor measurements (length and width) of mice
  
<u>Hypothesis to Test:</u>  
-The Combination of HDACi and PD1 antgaonist will have superior anti-tumor efficacy to single agents.  

```{r}
#| echo: false
#| warning: false
#| error: false
#| output: false

remove.packages("mouseExperiment")
devtools::install_github("https://github.com/SciOmicsLab/mouseExperiment")
#install.packages('/Users/david/Library/Mobile Documents/com~apple~CloudDocs/Programming/mouseExperiment', repos = NULL, type = "Source")

```


```{r libraries}
#| echo: false
#| warning: false
#| error: false

library(mouseExperiment)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(gt)
library(wesanderson)

```

```{r color-palette}
#| echo: false
#| warning: false
#| error: false

palette = wes_palettes$Darjeeling1

```

```{r load-data}
#| echo: false
#| warning: false
#| error: false

combo_experiment = read.csv("./data/combo_treatment_synthetic_data.csv")

treatment_data = read.csv("./data/combo_treatment_schedule.csv")

```

```{r add-day-volume}
#| echo: false
#| warning: false
#| error: false

combo_experiment = combo_experiment |> 
  calculate_dates(start_date = "03/17/2025") |> 
  calculate_volume()

treatment_data = treatment_data |> 
  calculate_dates(start_date = "03/17/2025")

```

# Summary Data

## Table 1. Sample Size Summary
```{r sample-size-summary}
#| echo: false

combo_experiment |> 
  dplyr::group_by(Treatment) |> 
  summarise(
    `Mice per Cage` = n_distinct(ID), 
    `Number of Cages` = n_distinct(Cage),
    `Total Sample Size` = `Mice per Cage` * `Number of Cages`
  ) |> 
  ungroup() |> 
  gt() |> 
  tab_header(title = md("**Sample Size Summary**"),
             subtitle = md("**Mice Per Group**")) |> 
  tab_style(
    style = cell_text(weight = "bold"),  # Make column names bold
    locations = cells_column_labels()  # Apply to column headers
  ) |> 
  tab_style(
    style = list(cell_fill(color = "lightgrey")), 
    locations = cells_body(rows = seq(1, 4, by = 2)) 
  ) |> 
  tab_style(
    style = cell_text(align = "center"),  # Center-align all text
    locations = cells_body()  # Apply to the entire body of the table
  )



```
Eight mice per treatment group, with four mice per cage were treated.


## Figure 1. Tumor Growth Plots
```{r tumor-growth-plots}
#| echo: false
#| fig-dpi: 200
#| fig-height: 10
#| fig-width: 7

p1 = plot_treatments(treatment_data = treatment_data,
                tumor_growth_data = combo_experiment,
                day_column = "Day",
                treatment_column = "Treatment",
                treatment_order = c("aPD1", "HDACi", "Control"),
                color_palette = palette[c(1,2,5)])

p2 = combo_experiment |>
  plot_tumor_growth(volume_column = "Volume", 
                    day_column = "Day", 
                    treatment_column = "Treatment", 
                    cage_column = "Cage", 
                    ID_column = "ID", 
                    survival_column = "Survival_Censor", 
                    extrapolate_volumes = FALSE,
                    group_summary_line = FALSE) +
    scale_color_manual(values = palette[c(5,1,2,3)])

p3 = combo_experiment |>
  plot_tumor_growth(volume_column = "Volume", 
                    day_column = "Day", 
                    treatment_column = "Treatment", 
                    cage_column = "Cage", 
                    ID_column = "ID", 
                    survival_column = "Survival_Censor", 
                    extrapolate_volumes = TRUE,
                    extrapolation_points = 3,
                    group_summary_line = TRUE) +
    scale_color_manual(values = palette[c(5,1,2,3)]) +
    ggtitle("Tumor Growth by Treatment and Dose\nwith Data Extrapolation")

ggarrange(p1,
          p2,
          p3,
          labels = "AUTO",
          nrow = 3,
          ncol = 1)

#clean up
rm(list = c("p1", "p2", "p3"))


```
Panel A: The treatment schedule of the indicated drugs is shown. Days of treatment for the corresponding treatment are indicated with a triangle. Panels B & C: Tumor volumes were estimated by the formula V = (length × width² × π) / 6. Tumor volume over time is plotted. Thin lines represent individual mice, colored by the treatment group as indicated in the legends. Thick lines are treatment group means (panel C). Panel B only includes data for surviving mice. Plot C extrapolates tumor growth curves for deceased mice using the last three available time-points/tumor measurements. 

# Tumor Growth Statistical Analysis

The "HDACi + PD1" group does not follow an exponential growth, so an area under the curve (auc) analysis was performed. 

```{r tumor-statistics}
#| echo: false
#| output: false

growth_statistics = tumor_growth_statistics(combo_experiment,
                                     time_column = "Day",
                                     volume_column = "Volume",
                                     treatment_column = "Treatment",
                                     cage_column = "Cage",
                                     id_column = "ID",
                                     model_type = "auc",
                                     extrapolation_points = 3,
                                     reference_group = "Control",
                                     handle_cage_effects = "as_random_effect")


```

```{r}
#| output: false
#| echo: false

growth_statistics$model
growth_statistics$anova
growth_statistics$summary
growth_statistics$posthoc
growth_statistics$treatment_effects
growth_statistics$growth_rates
growth_statistics$cage_analysis
growth_statistics$auc_analysis
growth_statistics$data_summary
growth_statistics$diagnostics


```


## Tumor Growth Rates
```{r growth-rates-plot}
#| echo: false
#| output: false
#| fig-dpi: 200
#| fig-height: 5
#| fig-width: 7

plot_growth_rate(growth_statistics$growth_rates,
         colors = palette[c(1,2,5,3)],
         group_order = c("Control", "HDACi", "aPD1", "HDACi + PD1")
         ) +
  theme_classic()


```

## Figure 2. Area Under the Curve.
```{r auc-plot}
#| echo: false
#| fig-dpi: 200
#| fig-height: 5
#| fig-width: 7

plot_auc(growth_statistics$auc_analysis$individual,
         show_mean = TRUE,
         error_bar_type = "CI",
         colors = palette[c(1,2,5,3)],
         group_order = c("Control", "HDACi", "aPD1", "HDACi + PD1"),
         extrapolated_column = "Extrapolated"
         ) +
  theme_classic()

```

Area under the curve for tumor volumes over time was calculated using the trapezoidal method for each mouse. Missing tumor measurements from deceased mice were calculated by extrapolation using a exponential growth model fitted with the three final measurement points. 

## Table 2. ANOVA Results 
```{r anova-results}
#| echo: false

growth_statistics$anova

```

## Table 3. Posthoc Comparisons 
```{r posthoc-results}
#| echo: false
#| warning: false

growth_statistics$posthoc$pairwise |>
  gt() |>
  cols_label(comparison = "Comparison",
             estimate = "Estimate",
             t_value = "t-Value",
             df = "Degrees of Freedom",
             ci_lower = "Lower 95% Confidence Interval",
             ci_upper = "Upper 95% Confidence Interval",
             p_value = "p-Value",
             p_adjusted = "Adjusted p-Value") |> 
  fmt_number(columns = c(p_value, p_adjusted), decimals = 4) |> 
  fmt_number(columns = c(estimate, ci_lower, ci_upper), decimals = 0) |> 
  fmt_number(columns = c(t_value, df), decimals = 2) |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = vars(p_adjusted),rows = p_adjusted < 0.05)) |> 
  tab_header(title = md("**Posthoc Pair-Wise Comparisons**"),
             subtitle = md("**Welch's t-test with Bonferroni Correction**")) |>
  tab_style(
    style = cell_text(weight = "bold"),  # Make column names bold
    locations = cells_column_labels()  # Apply to column headers
  ) |>
  tab_style(
    style = cell_text(align = "center"),  # Center-align all text
    locations = cells_body()  # Apply to the entire body of the table
  ) |> 
  tab_style(
    style = list(cell_fill(color = "lightgrey")),
    locations = cells_body(rows = seq(1, 6, by = 2))
  )


```
Statistical comparisons of area under the curve values were made by a type I ANOVA and Welch's t-tests with Bonferroni correction for multiple comparisons to account for the unequal variance between groups.

```{r power-analysis}
#| output: false
#| echo: false

power_analysis = post_power_analysis(combo_experiment,
                                     method = "auc")

power_analysis$effect_sizes
power_analysis$post_power_analysis
power_analysis$sample_size_estimates
power_analysis$summary
power_analysis$plots

#The post_power_analysis function returns the effect_sizes object with comparisons of some of the treatment groups, but not all possible group comparisons. It appears the reference column is only using one of the treatment groups. 

#The function should return a details object that includes information on how...

#make plotting functions?

```


# Drug Synergy

```{r synergy}
#| echo: false
#| output: false

synergy_stats = analyze_drug_synergy_over_time(combo_experiment,
                                               treatment_column = "Treatment",
                                               volume_column = "Volume",
                                               time_column = "Day",
                                               drug_a_name = "HDACi",
                                               drug_b_name = "aPD1",
                                               combo_name = "HDACi + PD1",
                                               control_name = "Control")

```

## Figure 4. Bliss Independence
```{r bliss-plot}
#| echo: false
#| fig-dpi: 200
#| fig-height: 5
#| fig-width: 8
#| warning: false

synergy_stats$synergy_summary |> 
  ggplot(aes(x = Time_Point)) +
    geom_line(aes(y = TGI_Drug_A, color = "Drug A")) +
    geom_line(aes(y = TGI_Drug_B, color = "Drug B")) +
    geom_line(aes(y = TGI_Combo, color = "Combination")) +
    geom_line(aes(y = Bliss_Expected_TGI, color = "Bliss Value"), alpha = 0.5, linewidth = 3) +
    scale_color_manual(values = c("Drug A" = "blue", 
                                  "Drug B" = "red", 
                                  "Combination" = "purple", 
                                  "Bliss Value" = "grey")) +
    labs(x = "Days", y = "Percent Tumor Growth Inhibition", color = NULL) +
    ylim(0,100) +
    theme_classic() +
    theme(legend.position = c(0.85,0.2))

#what is Drug A and Drug B? 

```
The percent tumor growth inhibition was calculated by the formula (1 - mean tumor volume of treatment group/mean tumor volume of control group) * 100 at each time point. The expected Bliss value, the percent tumor growth inhibition expected when two drugs act independently, is represented by the grey line. Percent tumor growth inhibition values for the drug combination above this curve indicate synergy; values within this curve represent additive effects, and values below this curve indicate antagonistic effects. 

## Figure 5. Combination Index
```{r combination-index-plot}
#| echo: false
#| fig-dpi: 200
#| fig-height: 4
#| fig-width: 6

synergy_stats$synergy_summary |> 
  ggplot(aes(x = Time_Point, y = Combination_Index))  +
    geom_line(color = "purple", linewidth = 1.5) +
    geom_ribbon(aes(ymin = 0, ymax = 0.85), fill = "blue", alpha = 0.2) +
    geom_ribbon(aes(ymin = 1.15, ymax = 2), fill = "red", alpha = 0.2) +
    geom_ribbon(aes(ymin = 0.85, ymax = 1.15), fill = "grey", alpha = 0.4) +
    annotate("text", x = mean(synergy_stats$synergy_summary$Time_Point), y = 0.2, label = "Synergistic", color = "blue", size = 5) +
    annotate("text", x = mean(synergy_stats$synergy_summary$Time_Point), y = 1.05, label = "Additive", size = 5) +
    annotate("text", x = mean(synergy_stats$synergy_summary$Time_Point), y = 1.8, label = "Antagonistic", color = "red", size = 5) + 
    labs(x = "Days", y = "Combination Index", title = "") +
    ylim(0,2) +
    theme_classic()


```
The combination index for the HDACi + anti-PD1 treatment group was calculated by dividing the tumor growth inhibition (TGI) of each drug by the tumor growth inhibition of the combination group and summing the values (TGI of HDACi/TGI of combo + TGI of anti-PD1/TGI of combo) at each time point. The plotted line represents the combination index over time. Values of less than 0.85 are considered synergistic, values between 0.85 to 1.15 are considered additive, and values above 1.15 are considered antagonistic. 


# Survival Analysis

## Figure 6. Survival Plot
```{r survival-plot}
#| echo: false
#| fig-dpi: 200
#| fig-height: 4
#| fig-width: 6

# plot_survival(combo_experiment,
#               time_column = "Day",
#               censor_column = "Survival_Censor",
#               treatment_column = "Treatment",
#               cage_column = "Cage",
#               id_column = "ID",
#               show_risk_table = TRUE,
#               title = "Survival Plot")

#Not working, remove for now

# Plot manually
combo_experiment |> 
  group_by(Treatment, Cage, ID) |> 
  slice_max(order_by = Day, n = 1, with_ties = FALSE) |> 
  ungroup() |> 
  survival::survfit(Surv(Day, Survival_Censor) ~ Treatment, data = _) |> 
  survminer::ggsurvplot(xlim = c(20,40), 
                        xlab = "Days",
                        risk.table = FALSE,
                        #risk.table.height = 0.4,
                        #risk.table.col = "black",
                        #risk.table.fontsize = 4,
                        legend = c(0.2, 0.25), 
                        legend.title = "", 
                        legend.labs = levels(combo_experiment$Treatment),
                        title = "Survival Plot",
                        palette = palette[c(5,1,2,3)], 
                        #theme = theme(strip.text = element_text(size = 4), risk.table.title = element_text(size = 4))
                      )
  
```
Kaplan-Meier curves for survival are shown. Lines are colored according to treatment groups. 


```{r survival-stats}
#| echo: false
#| output: false

survival_stats = survival_statistics(combo_experiment,
                                     time_column = "Day",
                                     censor_column = "Survival_Censor",
                                     treatment_column = "Treatment",
                                     cage_column = "Cage",
                                     id_column = "ID",
                                     reference_group = "Control")

```

## Table 4. Survival Statistics
```{r survival-table}
#| echo: false
#| warning: false


survival_stats$results |> 
  dplyr::select(-c("Median_Survival", "Event_Rate")) |> 
  gt() |> 
  cols_label(HR = "Hazard Ratio",
             CI_Lower = "Lower 95% Confidence Interval",
             CI_Upper = "Upper 95% Confidence Interval",
             P_Value = "p-Value",
             Total = "Sample Size") |> 
  fmt_number(columns = c(P_Value), decimals = 4) |> 
  fmt_number(columns = c(HR, CI_Lower, CI_Upper), decimals = 3) |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = vars(P_Value),rows = P_Value < 0.05)) |> 
  tab_header(title = md("**Cox Regression with Firth's Penalized Likelihood**")) |> 
  tab_style(
    style = cell_text(weight = "bold"),  # Make column names bold
    locations = cells_column_labels()  # Apply to column headers
  ) |>
  tab_style(
    style = cell_text(align = "center"),  # Center-align all text
    locations = cells_body()  # Apply to the entire body of the table
  ) |> 
  tab_style(
    style = list(cell_fill(color = "lightgrey")),
    locations = cells_body(rows = seq(1, 4, by = 2))
  )

```
Due to the low sample numbers, a Cox regression with Firth's penalized likelihood test was performed contrasting each treatment group against the "Control" group. 


```{r survival-results}
#| output: false
#| echo: false

survival_stats$results
survival_stats$reference_group
survival_stats$survival_data
survival_stats$method_used
survival_stats$model
survival_stats$km_plot

#km_plot isn't functioning properly. Remove? 

```

## Figure 7. Hazard Ratios
```{r forest-plot}
#| echo: false
#| fig-dpi: 200
#| fig-height: 4
#| fig-width: 8

# plot_forest(survival_stats$results,
#             group_order = c("Control", "HDACi", "aPD1", "HDACi + PD1"))

#The plot_forest function is not plotting correctly. Remove for now. 

#plot manually
ggplot(survival_stats$results, aes(y = factor(Group, levels = c("Control", "HDACi", "aPD1", "HDACi + PD1")), x = HR)) +
  geom_point(size = 5) +
  geom_errorbar(aes(xmin = CI_Lower, xmax = CI_Upper), width = 0.2) +
  geom_text(aes(label = paste0("HR: ", round(HR, 2), ", p = ", round(P_Value, 4))), vjust = 3, hjust = -0.2, size = 4) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  xlim(0, 2) +
  labs(x = "Hazard Ratio", y = "Treatment Group", title = "Survival Hazard Ratios") +
  theme_classic() +
  theme(axis.text.y = element_text(size = 11))

#remove plot_forest? 

```
The hazard ratios and 95% confidence intervals are plotted for each treatment group relative to the Control group. Hazard ratios less than 1 indicate a reduced risk of death (i.e. greater survival). 

# Conclusions
- Both the HDACi and anti-PD1 treatment groups had statistically significant reduced tumor burden relative to the control group. 
- The combination of HDACi and anti-PD1 had statistically significant reduced tumor burden relative to both single treatment groups.
- Bliss Independence and Combination Index modeling showed that the combination of HDACi and anti-PD1 had synergestic effects in tumor growth inhibition.
- HDACi, anti-PD1, and the combination treatment groups had significantly increased overall survival relative to the control group. Comparisons between the treatment groups were not made as too few events (deaths) occurred. If these comparisons are of interest, a longer study design is needed.
- **Overall, these data support the hypothesis that the Combination of HDACi and PD1 antgaonist will have superior anti-tumor efficacy to single agents.**
