---
title: "Mouse Experiment Analysis Example"
format: html
editor: visual
---

```{r setup, message=FALSE, warning=FALSE}
library(mouseExperiment)
library(dplyr)
library(ggplot2)
```

## Introduction

This document demonstrates the functionality of the mouseExperiment R package for analyzing and visualizing mouse tumor growth data.

## Data Preparation

First, we'll load the synthetic dataset included with the package:

```{r load-data}
data(synthetic_data)
head(synthetic_data)
```

Now we'll calculate tumor volumes and experiment days:

```{r calculate}
# Calculate tumor volume
df <- calculate_volume(synthetic_data)

# Calculate experiment days from start date
df <- calculate_dates(df, start_date = "2022-01-01")

# Display the resulting data
head(df)
```

## Tumor Growth Visualization

Let's visualize the tumor growth over time by treatment group:

```{r plot-growth}
plot_tumor_growth(df)
```

## Tumor Growth Statistics

We can perform statistical analysis on the tumor growth data:

```{r growth-stats, message=FALSE, warning=FALSE}
# Standard tumor growth statistics
growth_stats <- tumor_growth_statistics(df)
print(growth_stats)
```

## Survival Analysis

Now let's analyze and visualize survival data:

```{r survival-plot}
# Plot survival curves by group
plot_survival(df)
```

```{r survival-stats}
# Perform survival statistics
surv_stats <- survival_statistics(df)
print(surv_stats$results)
```

## Bayesian Analysis

The package also supports Bayesian analysis of tumor growth data:

```{r bayes-analysis, eval=FALSE}
# Bayesian tumor growth statistics 
# Note: This is computationally intensive, so we're not running it in this example
# bayesian_stats <- tumor_growth_statistics_bayes(df, cores = 2)
# print(bayesian_stats)
```

## Conclusion

This vignette demonstrates the key functionality of the mouseExperiment package for analyzing mouse tumor growth experiment data.